Last glacial cycle glacier erosion potential in the Alps

### 2021-05-20 articles/erosion/response-referee1.md

> Line 127 Higher precipitation increases ice flux and thus erosion, I guess?
> Would eventually helpful to mention (even though it might be referred in
> Seguinot et al., 2018).

Indeed. The increase in ice discharge was made explicit.

> Line 140 Can you be more precise what you mean by realistic? You mean because
> of the localized pattern? Maybe I am wrong but shouldn’t the erosion
> potential at least at the lake Constance overdeepening (Fig. 7) not in the
> order of hundreds of meters (cf. Preusser et al., 2010) and the best fit
> rather (b) or (c) – at least from what I read from the transect (e-h)?
> Visually (a)-(d) seem not to correspond to (e)-(h) if (a)-(d) is also
> presented in meters (annotation at the bar is missing). For example, in (b)
> the maximum erosion potential is like >>1000 m (if in meters) while in
> corresponding (f) it seems clearly lower than 1000 m.

TODO: "what you mean by realistic"

There is no mismatch between panels (a-d) and (e-h), but to improve readability
we have reduced the number of colour levels on the maps, and changed the
positions of ticks on the transects.

### 2021-05-22 articles/erosion/alps-erosion.tex
Add technical corrections from reviewer #2.

    The glacial erosion landscape of the Alps has fascinated generations of
    explorers, artists, mountaineers and scientists for centuries. Its cultural
    impact is indeed so far-reaching, that in English, a non-Alpine language,
    the adjective ``alpine'' with non-capital ``a'' is now casually used to
    describe an Alpine-like, glacially modified mountain landscape outside the
    Alps, while the proper noun ``Alps'' has been applied to nick-name
    Alpine-like, glacier eroded mountain ranges in Norway (\emph{Lyngsalpene}),
    New Zealand (Southern Alps), Japan
    (\emph{\begin{CJK}{UTF8}{min}日本アルプス\end{CJK},
    Nihon Arupusu}), and elsewhere.
    %
    Some mountain ranges are predominantly characterised by cirque glaciation
    (e.g. Uinta Mountains), glacial valleys (e.g. Putorana Plateau), or
    large-scale overdeepenings (e.g. Patagonia). But other regions, including
    the Alps \citep[e.g.][Fig.~\ref{fig:landscape}]{Penck.1905}, present a
    higher variety of glacier erosional landforms, whose implications on
    glacial history is yet to be understood.
    The cosmogenic nuclide memory of bedrock erosion provides a more
    quantitative, but equally varied picture of glacier erosion effectiveness
    both within and between glaciated regions \citep{Jansen.etal.2019,
    Steinemann.etal.2020, Steinemann.etal.2021}.

    In absence of other landforms, glacially eroded topography has sometimes
    been used as a proxy for glacier extent, for instance in
    geomorphological mapping \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
    Cold-based glaciers, however, have been observed to preserve landforms as
    fragile as sand beaches \citep{Kleman.1994}, leading to the
    reinterpretation of complex glacial landscapes within a binary conceptual
    framework of cold and temperate basal thermal regimes
    \citep{Kleman.etal.2008, Kleman.etal.2010, Fabel.etal.2012, Fu.etal.2013}.
    Nevertheless, it remains unclear how sharp the erosional
    transition to cold-based ice is, and why some glaciated regions show
    preserved periglacial blockfields and others, including the Alps, do not
    \citep{Wirsig.etal.2016, Seguinot.etal.2018}.

    Glacier erosion also likely governs, at least in part,
    the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through
    the ``glacier buzz-saw''. In turn, evidence suggests that increased
    mountain erosion rates coincided with global cooling and Pleistocene
    glaciations from roughly 2.6 million years before present
    \citep{Herman.Champagnac.2016}. However, oceanic isotopic proxies for
    global weathering rates remained constant over this time period
    \citep{Willenbring.Von-Blanckenburg.2010}. Examination of contemporary
    erosion rates across different climates suggest that increased temperatures
    in glaciated regions lead to higher erosion rates and that erosion rates
    may have increased towards present \citep{Koppes.Montgomery.2009,
    Koppes.etal.2015, Fernandez.etal.2016}. However, understanding the link
    between climate and glacier erosion in the sedimentary record remains
    difficult due to time scale biases \citep{Ganti.etal.2016} and the
    non-linear relationship  between atmospheric temperature and erosion
    \citep[e.g.,][]{Anderson.etal.2012,Mariotti.etal.2021}.

@@

    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
    previously published model results \citep{Seguinot.etal.2018} and the
    patterns of modelled erosion rates and cumulative last glacial cycle erosion
    potential are analysed. We examine the glaciological conditions that cause
    erosion and discuss the implications of these conditions on understanding
    the relationship between climate and glacier erosion. Despite aggregated
    uncertainties on paleoclimate, glacier flow and glacier erosion processes,
    out results provide insights into the diversity of the Alpine glacial
    erosion landscape.

@@

    The ice-sheet model set-up was presented in an earlier publication
    \citep{Seguinot.etal.2018} and is briefly summarized here. The simulations
    were conducted using the Parallel Ice Sheet Model (PISM, development
    version~e9d2d1f), an open-source, finite-difference, shallow-ice and
    shallow-shelf model \citep{PISM-authors.2017}. Our setup includes
    temperature and water-content dependent creep, pseudo-plastic basal
    sliding, bedrock deformation under the ice
    load, and a~positive degree-day (PDD) surface mass balance model. The model
    is initialized with assumed present-day ice thickness and equilibrium
    ice and bedrock temperature at 120\,ka (thousand years before the present),
    and ran to the present. The ice-sheet model physical parameters are listed
    in the earlier open-access publication \citep{Seguinot.etal.2018}, and the
    complete list of PISM
    parameters is stored in long-term archived model output metadata for the
    present study \citep{Seguinot.2021}, and for the earlier publication
    \citep{Seguinot.2020, Seguinot.2020a}.

    Climate forcing is provided by a~monthly climatology from interpolated
    observational data \citep[WorldClim;][]{Hijmans.etal.2005} and the European
    Centre for Medium-Range Weather Forecasts Reanalysis Interim
    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature lapse-rate
    corrections from the European Project for Ice Coring in Antarctica
    \citep[EPICA;][] {Jouzel.etal.2007}, and time-dependent paleo-precipitation
    reductions \citep{Huybrechts.2002}. Lower-resolution runs based on
    alternative paleoclimate scenarios \citep{Seguinot.etal.2018} are analyzed
    in the discussion part (Sect.~\ref{sec:sensitivity}).

@@

    A closer look at the Rhine Glacier, one of the paleo-ice sheet's largest
    outlets, reveals a spatial migration of the modelled rapid erosion areas.
    During stages of glacier advance and maximum extension, erosion is modelled
    to be under 1\,mm\,a$^{-1}$ and restricted the lower parts of the
    catchment, while much of the intra-montane Alps \citep[modelled to be
    largely cold-based, Fig.~6c in][]{Seguinot.etal.2018}, experience
    insignificant erosion (Fig.~\ref{fig:transects}a and~b). The modelled
    erosion rates both increase and propagate inwards during periods of glacier
    retreat (Fig.~\ref{fig:transects}c and~e).

@@

    The modelled, domain-integrated glacier erosion volume do not increase
    together with ice volume. On the opposite, periods of ice advance and
    maximum glaciation correspond to comparatively slow erosion
    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
    This result may seem paradoxical, and is modulated in a limited extent by
    the exponent, $l$, in the erosion rule (Fig.~\ref{fig:powerlaws}).
    Nevertheless, it may explain why some areas covered during glacial maxima
    only, such as some valleys on the southern side of the Alps, appear to have
    experienced only little glacial modification of their topography.
    During deglaciation periods, increased surface melt at low elevations
    results in a steepened topographic profile (Figs.~\ref{fig:basaldrag}a),
    as observed and further expected on currently retreating glaciers
    \citep{Huss.etal.2010, Zekollari.Huybrechts.2015}.
    The buttress formed by the glacier foot against gravitational forces is
    reduced, causing an increase in the magnitude of the basal shear stress
    (basal drag) relative to the ice overburden pressure, and an increase in
    the area where sliding and thus, erosion, can occur
    (Figs.~\ref{fig:basaldrag}b).

@@

    For instance, low-elevation piedmont over-deepened basins such as Lake
    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore, are
    here modelled to be the product of extensive, Last Glacial Maximum-like
    glaciations. Other modeling work suggests that features such as these
    likely form when water can access the glacier bed and encourage glacier
    sliding and, thus, erosion \citep{Herman.etal.2011}. The model herein does
    not include subglacial hydrology, yet the role of climate warming to
    increasing glacier erosion to the lower elevations remains evident
    (Fig.~\ref{fig:hypsogram}).
    Alpine terrain with elevations above 1000\,m are modelled to
    experience virtually no erosion from 35 to 18\,ka BP
    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
    intermediate-elevation glacial valleys of the Alps as Lauterbrunnental
    (Fig.~\ref{fig:landscape}), Haslital (Upper Aare) or Bout du Monde (Upper
    Giffre) would date from periods of intermediate glacier extension.
    The validity of the model results at high elevation are limited by the 1-km
    resolution, the absence of glacier cirque erosion processes
    \citep[cf.][]{Sanders.etal.2012}, and the the shallow-ice physics
    \citep{Imhof.etal.2019}.
    the lack of rapid erosion at high elevations for much
    of the modelled glacial cycle implies that the highest mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would have formed
    over interglacial periods, when glaciers are confined to high terrain
    \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
    Himalayan glaciers and may limit erosion for the high elevation portion of
    that mountain range during cold periods as well, failing to offset uplift
    rates \citep{Harper.Humphrey.2003}.

>>>>

    The glacial erosion landscape of the Alps has fascinated generations of
    explorers, artists, mountaineers and scientists for centuries. Its cultural
    impact is indeed so far-reaching, that in English, a non-Alpine language,
    the adjective ``alpine'' with non-capital ``a'' is now casually used to
    describe an Alpine-like, glacially modified mountain landscape outside the
    Alps, while the proper noun ``Alps'' has been applied to nick-name
    Alpine-like, glacier eroded mountain ranges in Norway (\emph{Lyngsalpene}),
    New Zealand (Southern Alps), Japan
    (\emph{\begin{CJK}{UTF8}{min}日本アルプス\end{CJK},
    Nihon Arupusu}), and elsewhere.
    %
    Some mountain ranges are predominantly characterised by cirque glaciation
    (e.g. Uinta Mountains), glacial valleys (e.g. Putorana Plateau), or
    large-scale overdeepenings (e.g. Patagonia). But other regions, including
    the Alps \citep[e.g.][Fig.~\ref{fig:landscape}]{Penck.1905}, present a
    higher variety of glacier erosional landforms, whose implications on
    glacial history are yet to be understood.
    The cosmogenic nuclide memory of bedrock erosion provides a more
    quantitative, but equally varied picture of glacier erosion effectiveness
    both within and between glaciated regions \citep{Jansen.etal.2019,
    Steinemann.etal.2020, Steinemann.etal.2021}.

    In the absence of other landforms, glacially eroded topography has
    sometimes been used as a proxy for glacier extent, for instance in
    geomorphological mapping \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
    Cold-based glaciers, however, have been observed to preserve landforms as
    fragile as sand beaches \citep{Kleman.1994}, leading to the
    reinterpretation of complex glacial landscapes within a binary conceptual
    framework of cold and temperate basal thermal regimes
    \citep{Kleman.etal.2008, Kleman.etal.2010, Fabel.etal.2012, Fu.etal.2013}.
    Nevertheless, it remains unclear how sharp the erosional
    transition to cold-based ice is, and why some glaciated regions show
    preserved periglacial blockfields and others, including the Alps, do not
    \citep{Wirsig.etal.2016, Seguinot.etal.2018}.

    Glacier erosion also likely governs, at least in part,
    the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through
    the ``glacier buzz-saw''. In turn, evidence suggests that increased
    mountain erosion rates coincided with global cooling and Pleistocene
    glaciations from roughly 2.6 million years before present
    \citep{Herman.Champagnac.2016}. However, oceanic isotopic proxies for
    global weathering rates remained constant over this time period
    \citep{Willenbring.Von-Blanckenburg.2010}. Examination of contemporary
    erosion rates across different climates suggests that increased temperatures
    in glaciated regions lead to higher erosion rates and that erosion rates
    may have increased towards present \citep{Koppes.Montgomery.2009,
    Koppes.etal.2015, Fernandez.etal.2016}. However, understanding the link
    between climate and glacier erosion in the sedimentary record remains
    difficult due to time scale biases \citep{Ganti.etal.2016} and the
    non-linear relationship  between atmospheric temperature and erosion
    \citep[e.g.,][]{Anderson.etal.2012,Mariotti.etal.2021}.

@@

    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
    previously published model results \citep{Seguinot.etal.2018} and the
    patterns of modelled erosion potential and cumulative last glacial cycle
    erosion
    potential are analysed. We examine the glaciological conditions that cause
    erosion and discuss the implications of these conditions on understanding
    the relationship between climate and glacier erosion. Despite aggregated
    uncertainties on paleoclimate, glacier flow and glacier erosion processes,
    out results provide insights into the diversity of the Alpine glacial
    erosion landscape.

@@

    The ice-sheet model set-up was presented in an earlier publication
    \citep{Seguinot.etal.2018} and is briefly summarized here. The simulations
    were conducted using the Parallel Ice Sheet Model (PISM, development
    version~e9d2d1f), an open-source, finite-difference, shallow-ice and
    shallow-shelf model \citep{PISM-authors.2017}. Our setup includes
    temperature and water-content dependent creep, pseudo-plastic basal
    sliding, bedrock isostatic deformation under the ice
    load, and a~positive degree-day (PDD) surface mass balance model. The model
    is initialized with assumed present-day ice thickness and equilibrium
    ice and bedrock temperature at 120\,ka (thousand years before the present),
    and is run to the present. The ice-sheet model physical parameters are listed
    in the earlier open-access publication \citep{Seguinot.etal.2018}, and the
    complete list of PISM
    parameters is stored in long-term archived model output metadata for the
    present study \citep{Seguinot.2021}, and for the earlier publication
    \citep{Seguinot.2020, Seguinot.2020a}.

    Climate forcing is provided by a~monthly climatology from interpolated
    observational data \citep[WorldClim;][]{Hijmans.etal.2005} and the European
    Centre for Medium-Range Weather Forecasts Reanalysis Interim
    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature changes
    from the European Project for Ice Coring in Antarctica
    \citep[EPICA;][] {Jouzel.etal.2007}, and time-dependent paleo-precipitation
    reductions \citep{Huybrechts.2002}. Lower-resolution runs based on
    alternative paleoclimate scenarios \citep{Seguinot.etal.2018} are analyzed
    in the discussion part (Sect.~\ref{sec:sensitivity}).

@@

    A closer look at the Rhine Glacier, one of the paleo-ice sheet's largest
    outlets, reveals a spatial migration of the modelled rapid erosion areas.
    During stages of glacier advance and maximum extension, erosion is modelled
    to be under 1\,mm\,a$^{-1}$ and restricted the lower parts of the
    catchment, while much of the intra-montane Alps \citep[modelled to be
    largely cold-based, Fig.~6c in][]{Seguinot.etal.2018} experience
    insignificant erosion (Fig.~\ref{fig:transects}a and~b). The modelled
    erosion rates both increase and propagate inwards during periods of glacier
    retreat (Fig.~\ref{fig:transects}c and~e).

@@

    The modelled, domain-integrated glacier erosion volume does not increase
    together with ice volume. On the opposite, periods of ice advance and
    maximum glaciation correspond to comparatively slow erosion
    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
    This result may seem paradoxical, and is modulated in a limited extent by
    the exponent, $l$, in the erosion rule (Fig.~\ref{fig:powerlaws}).
    Nevertheless, it may explain why some areas covered during glacial maxima
    only, such as some valleys on the southern side of the Alps, appear to have
    experienced only little glacial modification of their topography.
    During deglaciation periods, increased surface melt at low elevations
    results in a steepened topographic profile (Figs.~\ref{fig:basaldrag}a),
    as observed and further expected on currently retreating glaciers
    \citep{Huss.etal.2010, Zekollari.Huybrechts.2015}.
    The buttress formed by the glacier foot against gravitational forces is
    reduced, causing an increase in the magnitude of the basal shear stress
    (basal drag) relative to the ice overburden pressure, and an increase in
    the area where sliding and thus, erosion, can occur
    (Figs.~\ref{fig:basaldrag}b).

@@

    For instance, low-elevation piedmont over-deepened basins such as Lake
    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore are
    here modelled to be the product of extensive Last Glacial Maximum-like
    glaciations. Other modeling work suggests that features such as these
    likely form when water can access the glacier bed and encourage glacier
    sliding and, thus, erosion \citep{Herman.etal.2011}. The model herein does
    not include subglacial hydrology, yet the role of climate warming to
    increasing glacier erosion to the lower elevations remains evident
    (Fig.~\ref{fig:hypsogram}).
    Alpine terrain with elevations above 1000\,m are modelled to
    experience virtually no erosion from 35 to 18\,ka BP
    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
    intermediate-elevation glacial valleys of the Alps as Lauterbrunnental
    (Fig.~\ref{fig:landscape}), Haslital (Upper Aare) or Bout du Monde (Upper
    Giffre) would date from periods of intermediate glacier extension.
    The validity of the model results at high elevation are limited by the 1-km
    resolution, the absence of glacier cirque erosion processes
    \citep[cf.][]{Sanders.etal.2012}, and the the shallow-ice physics
    \citep{Imhof.etal.2019}.
    the lack of rapid erosion at high elevations for much
    of the modelled glacial cycle implies that the highest mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would have formed
    over interglacial periods, when glaciers are confined to high terrain
    \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
    Himalayan glaciers and may limit erosion for the high elevation portion of
    that mountain range during cold periods as well, failing to offset uplift
    rates \citep{Harper.Humphrey.2003}.
